---
title: "Advanced Abalone Analysis"
author: "mcgovern, kevin"
subtitle: "Data Analysis #2 (75 points total)"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
# library(rockchalk) # base R code replaces requirement for this package
library(car)
library(broom)
# color brewer
library(RColorBrewer)

# add library for table styling
library(kableExtra)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```


###Section 1 (5 points)
(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}

# build qq plot of ratio
q1.qqplot <- ggplot(data = mydata, aes(sample = RATIO)) +
  geom_qq(color='#2c7fb8') +
  geom_qq_line() +
  theme(legend.position = "none") +
  labs(x = "Theoretical", y = "Sample Ratio", title = "QQ Plot of Shuck Weight to Volume")

# build histogram of ratio
q1.hist <- ggplot(data = mydata, aes(x = RATIO)) +
  geom_histogram(binwidth=.02, fill='#2c7fb8') +
  theme(legend.position = "none") +
  labs(x = "Ratio", y = "Count", title = "Histogram of Shuck Weight to Volume")

grid.arrange(q1.hist, q1.qqplot, ncol=2)

# output skewness and kurtosis
print(paste("Skewness of ratio of shuck weight to volume", skewness(mydata$RATIO)))
print(paste("Kurtosis of ratio of shuck weight to volume", kurtosis(mydata$RATIO)))

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}
# calculate metric for log ratio
mydata$L_RATIO <- log10(mydata$RATIO)

# build qq plot of log ratio
q1.qqplot2 <- ggplot(data = mydata, aes(sample = L_RATIO)) +
  geom_qq(color='#2c7fb8') +
  geom_qq_line() +
  theme(legend.position = "none") +
  labs(x = "Theoretical", y = "Sample Ratio", title = "QQ Plot of Log10 Ratio")

# build histogram of log ratio
q1.hist2 <- ggplot(data = mydata, aes(x = L_RATIO)) +
  geom_histogram(binwidth=.02, fill='#2c7fb8') +
  theme(legend.position = "none") +
  labs(x = "Ratio", y = "Count", title = "Histogram of Log10 Ratio")

grid.arrange(q1.hist2, q1.qqplot2, ncol=2)

# create blue colors, excluding light colors
cb_blues = brewer.pal(n = 9, "Blues")[5:9]

# create boxplot of log ratio by class
ggplot(data = mydata, aes(y = L_RATIO, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_colour_manual(values = cb_blues) +
  labs(x = "Class", y = "Log of Ratio", title = "Log of Ratio by Class", color = "Class")

# output skewness and kurtosis
print(paste("Skewness of ratio of shuck weight to volume", skewness(mydata$L_RATIO)))
print(paste("Kurtosis of ratio of shuck weight to volume", kurtosis(mydata$L_RATIO)))
```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
# Bartlett test of homogeneity across classes using log ratio as dependent variable
print("Test log10 ratio homogeneity")
bartlett.test(mydata$L_RATIO ~ mydata$CLASS)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: The L_RATIO variable is much closer to a normal distribution than RATIO. The L_RATIO variable has skewness much closer to zero and kurtosis much closer to 3, both indicators of a normal distribution. Additionally, the histogram and the QQ-plot both display what appears to be much closer to a standard normal distribution for the L_RATIO variable. ***


###Section 2 (10 points)
(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

# Two Way ANOVA with interaction term CLASS:SEX
q2.aov_interaction <- aov(L_RATIO ~ CLASS*SEX, data = mydata)
summary(q2.aov_interaction)

# Two Way ANOVA without interaction term CLASS:SEX
q2.aov_wo_interaction <- aov(L_RATIO ~ CLASS + SEX, data = mydata)
summary(q2.aov_wo_interaction)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The non-significance interaction of Class and Sex as it relates to L_Ratio means that the values and effects of Class do not depend on Sex, and vice versa. Further, the pattern in the means of sex do not relate to the pattern in the means of class. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(q2.aov_wo_interaction, conf.level = 0.95)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: There is a significant difference in means for L_Ratio of all classes except A1-A2, so we can reject the null hypotheses of equal means in the difference in all class comparisons (except A1-A2) and accept the alternative hypotheses. For sex, we fail to reject the null that there is a significant difference between the means of males and females. As for the difference in means between infants and males, as well as infants and females, we reject the null hypothesis and accept alternative. This indicates that the interactions within the categories of class and sex are closely related in terms of how it influences L_Ratio for many of the classifications. Finally, the lack of significance in means of males and females means that we can feel comfortable combining the two sex categories into a single adult class with minimal risk.***


###Section 3 (10 points)
(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}

ggplot(data = mydata, aes(x = VOLUME, fill=TYPE)) +
  geom_histogram(bins = 15) +
  facet_grid(. ~ TYPE) +
  scale_fill_brewer(palette="Set2") +
  labs(x = "Volume", y = "Count", title = "Histogram of Abalone Volume")

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: Based on the histograms above, the distribution of infants appears to be skewed right with some outliers. The adult distribution appears to be much closer to a normal distribution. There is significantly more data captured for adults than there is for infants. There will definitely be overlap in infants and adults if volume is the sole field being used to separate the two types. The low values of infants make the smaller values easier to quantify for infants but the values above 100 volume appears to have a significant amount of overlap with adults. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}
# create log10 variables for shuck and volume
mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

# create a scatterplot of volume, split by class
q3.shuckVolScatter <- ggplot(data = mydata, aes(y = SHUCK, x = VOLUME, color = CLASS)) +
  geom_point() +
  scale_colour_manual(values = cb_blues) +
  labs(x = "Volume", y = "Shuck Weight (grams)", color = "Class")

# create a scatterplot of volume and Shuck Weight, split by class
q3.logShuckVolScatter <- ggplot(data = mydata, aes(y = L_SHUCK, x = L_VOLUME, color = CLASS)) +
  geom_point() +
  scale_colour_manual(values = cb_blues) +
  labs(x = "Log Volume", y = "Log Shuck Weight (grams)", color = "Class")

# create a scatterplot of volume and Shuck Weight, split by type
q3.shuckVolScatterType <- ggplot(data = mydata, aes(y = SHUCK, x = VOLUME, color = TYPE)) +
  geom_point() +
  scale_color_brewer(palette="Set2") +
  labs(x = "Volume", y = "Shuck Weight (grams)", color = "Type")

# create a scatterplot of Volume and Shuck Weight, split by type
q3.logShuckVolScatterType <- ggplot(data = mydata, aes(y = L_SHUCK, x = L_VOLUME, color = TYPE)) +
  geom_point() +
  scale_color_brewer(palette="Set2") +
  labs(x = "Log Volume", y = "Log Shuck Weight (grams)", color = "Type")

grid.arrange(q3.shuckVolScatter, q3.logShuckVolScatter, q3.shuckVolScatterType, q3.logShuckVolScatterType, ncol=2)

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: By applying the log on both sides of the scatterplot, it creates a much tighter band of data. This would very likely increase the Pearson correlation coefficient and make it closer to 1. Additionally the points move from clustered to the lower end of the range of both shuck weight and Volume, to the upper end of the chart for Log of Shuck Weight and Log of Volume. For linear regression, applying these log transformations, should reduce the error in the predictions of the chart. Class seems to be spread throughout the values but generally the lower class (A1, A2) tends to be at the lower end of the scatterplot, and the higher classes A4, A5 at the upper end. For type, in the chart without log transformations, adults and infants appear to overlap. When the volume and shuck weight have had log transformations applied, the infants separate a bit at the lower end, and the adults move to the upper portion of the chart. ***


###Section 4 (5 points)
(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))
```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

# create multiple linear regression
q4.linearReg <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
# output results
summary(q4.linearReg)

```

**Essay Question:  Interpret the trend in CLASS level estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The pattern in coefficients roughly is what we would expect where the upper level of classes provide more predictive power than the lower classes. This roughly aligns with what we've seen with the earlier TukeyHSD test where the difference in means of A1-A2 classes was the only non-significant difference. Finally, it is worth mentioning that the A2 class is just barely beyond the 0.1 level so depending on the level of alpha, it would still be important in some models. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: Type is not quite as predictive as some of the other classes, but it is still an important predictor in this model. With P values close to 0.006 it is still a very predictive indicator, even if it isn't quite to the level of some of the other class variables. Also, the fact that the R-Squared and the Adjusted R-Squared are so close together is an indication that there are not many non-important predictors in the regression model. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

### Section 5 (5 points)
(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}

model_data <- augment(q4.linearReg)

# build qq plot of residuals
q5.qqplot <- ggplot(data = model_data, aes(sample = .resid)) +
  geom_qq(color='#2c7fb8') +
  geom_qq_line() +
  theme(legend.position = "none") +
  labs(x = "Theoretical", y = "Sample Ratio", title = "QQ Plot of Shuck Weight to Volume")

# build histogram of residuals
q5.hist <- ggplot(data = model_data, aes(x = .resid)) +
  geom_histogram(bins = 20, fill='#2c7fb8') +
  theme(legend.position = "none") +
  labs(x = "Ratio", y = "Count", title = "Histogram of Shuck Weight to Volume")

grid.arrange(q5.hist, q5.qqplot, ncol=2)

# output skewness and kurtosis
print(paste("Skewness of ratio of shuck weight to volume", skewness(q4.linearReg$residuals)))
print(paste("Kurtosis of ratio of shuck weight to volume", kurtosis(q4.linearReg$residuals)))

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}

# create a scatterplot of residuals vs log of volume, colored by rings
q5.scatter_class <- ggplot(data = model_data, aes(y = .resid, x = L_VOLUME, color=CLASS)) +
  geom_point() +
  scale_colour_manual(values = cb_blues) +
  labs(y = "Residuals", x = "Log of Volume", color = "Class")

# create a scatterplot of residuals vs log of volume, colored by type
q5.scatter_type <- ggplot(data = model_data, aes(y = .resid, x = L_VOLUME, color=TYPE)) +
  geom_point() +
  scale_color_brewer(palette="Set2") +
  labs(y = "Residuals", x = "Log of Volume", color = "Type")

# create a boxplot of residuals, split by class
q5.boxplot_class <- ggplot(data = model_data, aes(y = .resid, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_colour_manual(values = cb_blues) +
  labs(x = "Class", y="Residuals", color = "Class")

# create a boxplot of residuals, split by type
q5.boxplot_type <- ggplot(data = model_data, aes(y = .resid, x = TYPE, color=TYPE)) +
  geom_boxplot() +
  scale_color_brewer(palette="Set2") +
  labs(x = "Type", y="Residuals", color = "Type")

grid.arrange(q5.scatter_class, q5.scatter_type, q5.boxplot_class, q5.boxplot_type, ncol=2)

# output bartlett test for residuals
bartlett.test(q4.linearReg$residuals ~ mydata$CLASS)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: With skewness near 0 and kurtosis near 3 of the residuals, the values appears to be normally distributed, which indicates a good fit. Additionally, in the scatterplot, the values are tightly packed towards the upper end of the log of volume but the residuals are still evenly spread above and below 0. There are limited outliers in both classes and types of residuals and the quartiles seem to be tightly packed. Overall, the model looks to fit the data well and L_Volume should be a good proxy for making decisions on. ***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

###Section 6 (5 points)
(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}

# create data frame to be used in ggplot
q6.df <- data.frame(volume.value, prop.adults, prop.infants)

q6.df$prop.adults.harvested <- 1 - prop.adults
q6.df$prop.infants.harvested <- 1 - prop.infants

# create plot of splits
ggplot(data = q6.df, aes(x = volume.value)) +
  geom_line(aes(y = prop.adults, color = "Adult")) +
  geom_line(aes(y = prop.infants, color = "Infant")) +
  geom_vline(xintercept = split.adults, linetype = "dashed", color = "#696969") +
  geom_vline(xintercept = split.infants, linetype = "dashed", color = "#696969") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#696969") +
  annotate("text", x = 430, y = 0.47, label = round(split.adults, digits = 2)) +
  annotate("text", x = 180, y = 0.47, label = round(split.infants, digits = 2)) +
  labs(x = "Volume", y="Proportions", title = "Infant vs. Adult Proportions as a Function of Volume") +
  scale_color_manual(name = "Proportions", values = c("#016450","#3690c0"))

```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: These cutoff points highlight the difference in medians of infants and adults. At a much lower cutoff, specifically anything above 133.82, we would harvest more than 50% of the infant population. Whereas, the cutoff point where we would harvest more than 50% of the adult population doesn't occur until the cutoff point exceeds 384.51. ***

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###Section 7 (10 points)
(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

# create vector of differences
differences <- ((1 - prop.adults) - (1 - prop.infants))

# create plot of splits
ggplot(data = q6.df, aes(x = volume.value, y = differences)) +
  geom_line(color = "#016450") +
  geom_vline(xintercept = split.adults, linetype = "dashed", color = "#696969") +
  geom_vline(xintercept = split.infants, linetype = "dashed", color = "#696969") +
  labs(x = "Volume", y="Difference in Proportions", title = "Difference in Infant and Adult Proportions of Volume")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}

# find max value of smoothed line of differences
q7.maxVal <- q6.df[which.max(smooth.difference), ]$volume.value

# create plot of splits
ggplot(data = q6.df, aes(x = volume.value)) +
  geom_line(aes(y = differences), color = "#016450") +
  geom_line(aes(y = smooth.difference), color = "#016450") +
  geom_vline(xintercept = q7.maxVal, linetype = "dashed", color = "#696969") +
  annotate("text", x = q7.maxVal + 50, y = 0.5, label = round(q7.maxVal, digits = 2)) +
  labs(x = "Volume", y="Difference in Proportions", title = "Difference in Infant and Adult Proportions of Volume with Smoothed Difference")

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
# compute differences
max.difference.adult <- (1 - prop.adults)[which.max(smooth.difference)]
max.difference.infant <- (1 - prop.infants)[which.max(smooth.difference)]

# Adult output
print(paste("Adult harvest proportion of smoothed max difference:", max.difference.adult)) # [1] 0.7416332
# Infant output
print(paste("Infant harvest proportion of smoothed max difference:", max.difference.infant))

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

###Section 8 (10 points)
(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
# find first volume variable where there are no A1 Infants
q8.cutoff <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

# calculate row number for above cutoff point
q8.cutoffRow <- which(volume.value == q8.cutoff)
# calculate adult and infant proportions
noA1Infants.adult <- (1 - prop.adults)[q8.cutoffRow]
noA1Infants.infant <- (1 - prop.infants)[q8.cutoffRow]

# Adult output
print(paste("Adult harvest proportion of no A1 infants method:", noA1Infants.adult))
# Infant output
print(paste("Infant harvest proportion of no A1 infants method:", noA1Infants.infant))

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
q8.equalRatesCutoff <- volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391

# calculate row number for above cutoff point
q8.equalRatesCutoffRow <- which(volume.value == q8.equalRatesCutoff)
# calculate adult and infant proportions
equalRates.adult <- (1 - prop.adults)[q8.equalRatesCutoffRow]
equalRates.infant <- (1 - prop.infants)[q8.equalRatesCutoffRow]

# Adult output
print(paste("Adult harvest proportion of equal errors method:", equalRates.adult))
# Infant output
print(paste("Infant harvest proportion of equal errors method:", equalRates.infant))

```


###Section 9 (5 points)
(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}
# create data frame of points
q9.pointlist <- list()
q9.pointlist[[1]] <- data.frame(infants = max.difference.infant, adults = max.difference.adult, 
                                text = "max difference", volume = volume.value[which.max(smooth.difference)],
                                yield = (max.difference.infant*total.infants + max.difference.adult*total.adults)/(total.adults+total.infants))
q9.pointlist[[2]] <- data.frame(infants = noA1Infants.infant, adults = noA1Infants.adult, 
                                text = "zero A1 infants", volume = q8.cutoff,
                                yield = (noA1Infants.infant*total.infants + noA1Infants.adult*total.adults)/(total.adults+total.infants))
q9.pointlist[[3]] <- data.frame(infants = equalRates.infant, adults = equalRates.adult, 
                                text = "equal errors", volume = q8.equalRatesCutoff,
                                yield = (equalRates.infant*total.infants + equalRates.adult*total.adults)/(total.adults+total.infants))
q9.points = do.call(rbind, q9.pointlist)


# create plot of splits
ggplot(data = q6.df, aes(y = prop.adults.harvested, x = prop.infants.harvested)) +
  geom_line(color = "#016450") +
  geom_point(data = q9.points, aes(x = infants, y = adults)) +
  geom_abline(slope = 1, linetype = "dashed", color = "#696969") +
  annotate("text", x = q9.points$infants + .18, y = q9.points$adults - .01, label = paste(q9.points$text, "volume:", round(q9.points$volume, digits = 2))) +
  labs(x = "Volume", y="Difference in Proportions", title = "Difference in Infant and Adult Proportions of Volume with Smoothed Difference")


```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
print(paste("The Area Under the Curve is", auc(q6.df$prop.infants.harvested, q6.df$prop.adults.harvested)))

```


###Section 10 (10 points)
(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	
# reorder and rename columns
q10.df <- q9.points[,c(3,4,2,1,5)]
colnames(q10.df) <- c("Cutoff Point", "Volume", "True Positive", "False Positive", "Proportion Yield")
# output data frame
kable(q10.df, format = "html", caption = "Cutoffs Method Summary") %>% kable_styling()

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: The three cutoff methods all produce cutoff points that occur at different volumes. The first cutoff point uses the location where the difference between the harvested percentages of adults and infants is maximized. The smoothing method allows this to occur closer to the midway point between the maximum not harvested percentages of adults and infants. This cutoff point occurs at the highest volume of the three methods evaluated. The max difference method also results in the lowest false positive rate, however it also results in the lowest true positive rate, leading it to be the most conservative method. The method that finds that point at which there are no more A1 infants results in the highest true positive rate, but also the highest false positive rate. Additionally, the proportion yield is the highest of the three methods. Finally, the cutoff point method that finds the location where the error rates of the adults and infants is equal is a middle of the road cutoff point that ranks second in the true positive rate, the false positive rate, and the proportion yield. Interestingly. Overall, we can see that for the cutoff points evaluated, as the volume of the cutoff point decreases, the true positive, false positive, and proportion yields all increase (negative correlation).   ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?  4)  What suggestions would you have for planning future abalone studies of this type? 

***Answer: A presentation of these findings should start out with a statement on the purpose of the study and the problem that this analysis would be looking to solve, similar to the introduction included in the first report. There should also be an outline of the overall problems encountered during the first and second data analysis assignments and discussion of the assumptions that were made. Additionally, the manipulations performed on the data to get it to a format where we can drive meaningful results should be noted. Some of the manipulations I would call out include taking the log of different values and creating a new combined category for sex. One of the limitations of the data that should be mentioned again is the discovery found in the first data analysis assignment that there was some confounding results in the class and sex variables. Additionally, the fact that the log of volume was used in the predictive model is another area that should be clear. As for suggestions to implement a cutoff, given how easy it would be to overharvest, the investigators should consider starting with a low cutoff and see how well that works for them. As with all things in data science, this is a balancing act and there are trade-offs for one method over another. The investigators will also need to decide if they value higher true positive rates or lower false positive rates more. Depending on which is more important, they can decide on a cutoff based on the values specified. For future studies, I would recommend to the researchers conducting the study that they use caution with classifications of abalone sex and age. The decisions made early on in the process will heavily influence the later analysis. A potential presentation would be concluded with a summary of high level findings from the analysis and additional appendices that cover the in-depth details of the reports. ***