---
title: 'Exploratory Abalone Analysis'
subtitle: 'Data Analysis Assignment #1 (50 points total)'
author: "McGovern, Kevin"
date: "February 10th, 2019"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

DESCRIPTION GOES HERE
Type of data, how much data, types of data, what is an abalone

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# load color libraries
library(colorspace)
library(RColorBrewer)

# add library for table styling
library(kableExtra)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

(1)(a) (2 points) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)

# create table with class and rings as dimensions
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Sex and Class are both categorical variables, Rings is an integer, all other variables are decimal place(Enter your answer here.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
SexByClass <- table(mydata$SEX, mydata$CLASS)

addmargins(SexByClass)

barplot(SexByClass, beside=T, col=brewer.pal(3, "Blues"), border="black",
     main = "Count of Abelones by Class, grouped by Sex",
     xlab="Class grouped by Sex", ylab="Count")

legend(x="topright", legend=c("Female","Infant","Male"),
       col=brewer.pal(3, "Blues"), pch=15)

```

**Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (Enter your answer here.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
myDataInts <- c(1:nrow(mydata))
sampleRows <- sample(myDataInts, 200)

work <- mydata[sampleRows,]

plot(work[, 2:6])
# use print commands
# use fig command for plot

```

-----

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$WHOLE, mydata$VOLUME, pch = 16, col=brewer.pal(5, "YlGnBu")[mydata$CLASS],
     main = "Comparison of Weight of Whole Abelone to Abelone Volume",
     xlab="Whole Weight (grams)", ylab="Volume (cm^3)")

legend(x="bottomright", legend=unique(mydata$CLASS),
       col=brewer.pal(5, "YlGnBu"), pch=16)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
# max value of ratio in shuck to whole


plot(mydata$WHOLE, mydata$SHUCK, pch = 16, col=rainbow_hcl(5)[mydata$CLASS],
     main = "Comparison of Weight of Whole to Shucked Weight of Abelone",
     xlab="Whole Weight (grams)", ylab="Shucked Weight (grams)")
# add color to this line
abline(0, max(mydata$SHUCK/mydata$WHOLE))
# add text describing meaning of line
#text(maxX, maxY-.1, paste("max value ",round(maxX, digits = 3), ",", round(maxY, digits = 3), sep = ""),
#     cex = .8)
legend(x="bottomright", legend=unique(mydata$CLASS),
       col=rainbow_hcl(5), pch=16)

```

**Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (Enter your answer here.)***

-----

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

# fix labels and add titles

ggHist <- ggplot(data = mydata, aes(x = RATIO, fill=SEX)) +
  geom_histogram(binwidth=.02) +
  facet_grid(. ~ SEX) +
  scale_fill_brewer(palette="Set2")

ggBloxplot <- ggplot(data = mydata, aes(y = RATIO, color=SEX)) +
  geom_boxplot(width=1) +
  xlim(-1, 1) +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2")

ggQQPlot <- ggplot(data = mydata, aes(sample = RATIO, color=SEX)) +
  geom_qq() +
  geom_qq_line() +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2")

grid.arrange(ggHist, ggBloxplot, ggQQPlot)
```

**Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Enter your answer here.)***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
# create a function to determine the mild outliers using a boxplot
findOutliers <- function(sex){
  return(boxplot.stats(x = mydata$RATIO[mydata$SEX == sex])[["out"]])
}
# create a function to determine extreme outliers using 3*IQR
findExtremeOutliers <- function(outliers, sex){
  return(outliers > (quantile(mydata$RATIO[mydata$SEX == sex], .75) + 3.0*IQR(mydata$RATIO[mydata$SEX == sex])))
}

# use outlier function to determine outliers for male, female, and infants
ratioOutliers_m <- findOutliers("M")
ratioOutliers_f <- findOutliers("F")
ratioOutliers_i <- findOutliers("I")

# get row positions of mild outliers
ratioOutliers_positions <- match(c(ratioOutliers_m, ratioOutliers_f, ratioOutliers_i), mydata$RATIO)

# output all mild outliers
kable(mydata[ratioOutliers_positions, ], format = "html", caption = "Mild Ratio Outliers Determined by Sex") %>% kable_styling()

# use extreme outlier function to determine extreme outliers for male, female, and infants
ratioOutliersExtreme_m <- findExtremeOutliers(ratioOutliers_m, "M")
ratioOutliersExtreme_f <- findExtremeOutliers(ratioOutliers_f, "F")
ratioOutliersExtreme_i <- findExtremeOutliers(ratioOutliers_i, "I")

# create vector of all extreme outliers
ratioOutliersExtreme <- c(ratioOutliers_m[ratioOutliersExtreme_m],ratioOutliers_f[ratioOutliersExtreme_f],ratioOutliers_i[ratioOutliersExtreme_i])
# find row positions of extreme outliers
ratioOutliersExtreme_positions <- match(ratioOutliersExtreme, mydata$RATIO)

# output all extreme outliers
kable(mydata[ratioOutliersExtreme_positions, ], format = "html", caption = "Extreme Ratio Outliers Determined by Sex") %>% kable_styling()


```

**Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (Enter your answer here.)***

-----

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
ggVolumeBoxplt <- ggplot(data = mydata, aes(y = VOLUME, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Class")
ggWholeBoxplt <- ggplot(data = mydata, aes(y = WHOLE, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Whole Weight (grams)", color = "Class")
ggVolumeScatter <- ggplot(data = mydata, aes(y = VOLUME, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Class")
ggWholeScatter <- ggplot(data = mydata, aes(y = WHOLE, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Whole Weight (grams)", color = "Class")

grid.arrange(ggVolumeBoxplt, ggWholeBoxplt, ggVolumeScatter, ggWholeScatter, ncol=2)

```

**Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: There is a general positive trend between the number of rings and both the Whole and Volume variables. However, both variables have a wide range of variability for the A3, A4, and A5 classes. Therefore, I believe that both volume and whole would not perform well as predictors for age, unless the age range is limited to very young abelones.***

-----

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
mydataAgg <- aggregate(x = mydata[,c("VOLUME","SHUCK","RATIO")], by = list(mydata$SEX, mydata$CLASS), FUN = "mean")

# create volume matrix
volumeMatrix <- matrix(mydataAgg$VOLUME, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))
# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Volume by Sex and Class") %>% kable_styling()

# create shuck matrix
shuckMatrix <- matrix(mydataAgg$SHUCK, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))
# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Shuck Weight by Sex and Class") %>% kable_styling()

# create ratio matrix
ratioMatrix <- matrix(mydataAgg$RATIO, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))

# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Ratio (Shuck Weight to Volume) by Sex and Class") %>% kable_styling()

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
# create ratio to class chart
ggplot(data = mydataAgg, aes(y = RATIO, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Ratio of Shuck Weight to Volume", color = "Sex", title = "Mean Ratio of Shuck Weight to Volume by Class")

# create volume to class chart
ggplot(data = mydataAgg, aes(y = VOLUME, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Sex", title = "Mean Volume by Class")

# create shuck to class chart
ggplot(data = mydataAgg, aes(y = SHUCK, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Shuck Weight (grams)", color = "Sex", title = "Mean Shuck Weight by Class")

```

**Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
mydataSubset_i <- subset(mydata, RINGS < 16 & SEX == 'I')
mydataSubset_a <- subset(mydata, RINGS < 16 & (SEX == 'M' | SEX == 'F'))


par(mfrow = c(2, 2))
boxplot(VOLUME ~ RINGS, data = mydataSubset_i, col = rgb(44, 127, 184, maxColorValue = 255), ylim = c(0, 1100),
     main = "Rings of infants by Volume", ylab = "Volume")
boxplot(VOLUME ~ RINGS, data = mydataSubset_a, col = rgb(127, 205, 187, maxColorValue = 255), ylim = c(0, 1100),
     main = "Rings of adults by Volume", ylab = "Volume")
boxplot(WHOLE ~ RINGS, data = mydataSubset_i, col = rgb(44, 127, 184, maxColorValue = 255), ylim = c(0, 400),
     main = "Rings of infants by Whole Weight", ylab = "Whole Weight")
boxplot(WHOLE ~ RINGS, data = mydataSubset_a, col = rgb(127, 205, 187, maxColorValue = 255), ylim = c(0, 400),
     main = "Rings of adults by Whole Weight", ylab = "Whole Weight")


```

**Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (Enter your answer here.)***

-----

**Conclusions**

Please respond to each of the following questions (10 points total):

**Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Enter your answer here.)***

**Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (Enter your answer here.)***

**Question 3)  (2 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Enter your answer here.)***
