---
title: 'Exploratory Abalone Analysis'
subtitle: 'Data Analysis Assignment #1 (50 points total)'
author: "McGovern, Kevin"
date: "February 10th, 2019"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

DESCRIPTION GOES HERE
Type of data, how much data, types of data, what is an abalone

Non-uniform class
Messy classes
shucked weight does not increase at same pace as whole weight in older abelone

1-2 sentences on conculsion of project

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# load color libraries
library(colorspace)
library(RColorBrewer)

# add library for table styling
library(kableExtra)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

(1)(a) (2 points) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)

# create table with class and rings as dimensions
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Sex is a categorical variables. There are 55 more males than females, so that could skew some of the results later to overrepresent male abelone. Length and diameter appear to be roughly normally distributed because the mean and median are close to equal but it looks like the left tail contains one or more an outliers (will have to see in further analysis). Whole weight and volume look to be normally distributed based on these statistics as well but the range is much larger and the mean is only marginally higher than the median. Shuck appears to have a near-normal interquartile range but with such a large max value, it is possible that there are outliers on the right tail. Rings is an integer and because it is a rough estimation of age, the expectation should be for a uniform distribution to make sure that all ages are represented equally, however the statistics here do not show that. The Class variable is another categorical variable that is a representation of rings and age. As you can see from the summary, there are a high proportion of A3 abelones. Ratio was a variable we calculated by taking shuck and dividing by volume. Because the two variables used to calculate the value are near normal, this ratio appears close to normal as well.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
SexByClass <- table(mydata$SEX, mydata$CLASS)

addmargins(SexByClass)

barplot(SexByClass, beside=T, col=brewer.pal(3, "Blues"), border="black",
     main = "Count of Abelones by Class, grouped by Sex",
     xlab="Class grouped by Sex", ylab="Count")

legend(x="topright", legend=c("Female","Infant","Male"),
       col=brewer.pal(3, "Blues"), pch=15)

```

**Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The class variable is a classification of ages based on rings, with A1 being the youngest. Some things stand out about the distribution here. The first observation is that because A1 is the youngest class of abelone, the expectation is that all abelone are classified as infant and there wouldn't be any male or female abelone in this class. The second thing is, that I would expect all of the infant abelone to be in either class A1 or A2 but I see a 60 or so infants in A3 and around 20 infants in the A4 and A5 classes. It seems as though these classifications were not created correctly otherwise, the male. The last thing is that in the A2 and A3 classes, there is a much higher proportion of males than females. This indicates that there is a possibility that the sample taken was either not random, or that male abelone are more represented in the younger population than females.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
myDataInts <- c(1:nrow(mydata))
sampleRows <- sample(myDataInts, 200)

work <- mydata[sampleRows,]

plot(work[, 2:6], pch = 16, , cex = 0.7, col=rgb(1, 108, 89, maxColorValue = 255))
# use print commands
# use fig command for plot

```

-----

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
# sort data so coloring will be applied correctly
mydata.sorted <- mydata[order(mydata$CLASS),] 

plot(mydata.sorted$WHOLE, mydata.sorted$VOLUME, pch = 16, col=brewer.pal(5, "YlGnBu")[mydata.sorted$CLASS],
     main = "Comparison of Weight of Whole Abelone to Abelone Volume",
     xlab="Whole Weight (grams)", ylab="Volume (cm^3)")

legend(x="bottomright", legend=unique(mydata.sorted$CLASS),
       col=brewer.pal(5, "YlGnBu"), pch=16)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}

# create basic scatter plot with whole weight on the x-axis and shuck on the y-axis
plot(mydata.sorted$WHOLE, mydata.sorted$SHUCK, pch = 16, col=brewer.pal(5, "YlGnBu")[mydata.sorted$CLASS],
     main = "Comparison of Weight of Whole to Shucked Weight of Abelone",
     xlab="Whole Weight (grams)", ylab="Shucked Weight (grams)")
# draw the line reprenting the maximum ratio of shuck to whole weight
abline(0, max(mydata.sorted$SHUCK/mydata.sorted$WHOLE), col=rgb(1, 108, 89, maxColorValue = 255))

# create text showing point of line intersection
maxShucktoWhole <- which.max(mydata.sorted$SHUCK/mydata.sorted$WHOLE)
text(mydata.sorted[maxShucktoWhole,]$WHOLE, mydata.sorted[maxShucktoWhole,]$SHUCK*1.08, 
  "max ratio", cex = .8)

# add legend indicating color meaning
legend(x="bottomright", legend=unique(mydata.sorted$CLASS),
       col=brewer.pal(5, "YlGnBu"), pch=16)

```

**Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: Both plots have a rough positive correlation and similar shape. The plots start out with a tight correlation at the lower values but become more dispersed as the values become larger. The chart comparing abelone volume to whole weight is more dispersed in the higher values than the chart comparing whole weight to shucked weight. In the volume-whole weight comparison chart, regardless of age class, the abelone appear to be spread somewhat evenly across the chart. However, in the whole weight to shucked weight comparison chart, it seems like whole weight continues to increase with age but shucked weight does not increase at the same pace. The older abelone appear to drop in their ratio of whole to shucked weight. Additionally, the maximum value of ratio between whole and shucked weight is in the A3 class, not in the A4 or A5 class.***

-----

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

# fix labels and add titles

ggHist <- ggplot(data = mydata, aes(x = RATIO, fill=SEX)) +
  geom_histogram(binwidth=.02) +
  facet_grid(. ~ SEX) +
  scale_fill_brewer(palette="Set2")

ggBloxplot <- ggplot(data = mydata, aes(y = RATIO, color=SEX)) +
  geom_boxplot(width=1) +
  xlim(-1, 1) +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2")

ggQQPlot <- ggplot(data = mydata, aes(sample = RATIO, color=SEX)) +
  geom_qq() +
  geom_qq_line() +
  facet_grid(. ~ SEX) +
  scale_color_brewer(palette="Set2")

grid.arrange(ggHist, ggBloxplot, ggQQPlot)
```

**Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: Each of the three charts above have a slight skew to the right. The histogram shows that there is not a mound shape to the data typically found in a normal distribution. However, the histograms are also not heavily skewed right. The boxplots show the values are tightly packed together around the mean but each has a few high outliers in the right part of the tail. The QQ Plots show that most of the points fall near the QQ line with the exception of the right tail of the data. There is one point in the female chart that is very far from the QQ line that as we'll see later is an extreme outlier.***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
# create a function to determine the mild outliers using a boxplot
findOutliers <- function(sex){
  return(boxplot.stats(x = mydata$RATIO[mydata$SEX == sex])[["out"]])
}
# create a function to determine extreme outliers using 3*IQR
findExtremeOutliers <- function(outliers, sex){
  return(outliers > (quantile(mydata$RATIO[mydata$SEX == sex], .75) + 3.0*IQR(mydata$RATIO[mydata$SEX == sex])))
}

# use outlier function to determine outliers for male, female, and infants
ratioOutliers_m <- findOutliers("M")
ratioOutliers_f <- findOutliers("F")
ratioOutliers_i <- findOutliers("I")

# get row positions of mild outliers
ratioOutliers_positions <- match(c(ratioOutliers_m, ratioOutliers_f, ratioOutliers_i), mydata$RATIO)

# output all mild outliers
kable(mydata[ratioOutliers_positions, ], format = "html", caption = "Mild Ratio Outliers Determined by Sex") %>% kable_styling()

# use extreme outlier function to determine extreme outliers for male, female, and infants
ratioOutliersExtreme_m <- findExtremeOutliers(ratioOutliers_m, "M")
ratioOutliersExtreme_f <- findExtremeOutliers(ratioOutliers_f, "F")
ratioOutliersExtreme_i <- findExtremeOutliers(ratioOutliers_i, "I")

# create vector of all extreme outliers
ratioOutliersExtreme <- c(ratioOutliers_m[ratioOutliersExtreme_m],ratioOutliers_f[ratioOutliersExtreme_f],ratioOutliers_i[ratioOutliersExtreme_i])
# find row positions of extreme outliers
ratioOutliersExtreme_positions <- match(ratioOutliersExtreme, mydata$RATIO)

# output all extreme outliers
kable(mydata[ratioOutliersExtreme_positions, ], format = "html", caption = "Extreme Ratio Outliers Determined by Sex") %>% kable_styling()


```

**Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There are quite a few mild outliers (19 to be exact). As we saw from the charts above, the plots are slightly skewed right so this is expected. There are also two extreme outliers. The female outlier was immiedately apparent when viewing the small multiple charts created prior to the outlier calculations so that was not shocking. The one infant outlier was just barely beyond the threshold of what constitutes an extreme outlier (.269 vs a threshold of .2685) which is why it was not as noticable as the female outlier. Overall, due to the skew, it is not surprising that there are so many mild outliers or that there is such a high proportion of mild outliers to extreme outliers because the distributions are either normal or extremely skewed.***

-----

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
ggVolumeBoxplt <- ggplot(data = mydata, aes(y = VOLUME, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Class")
ggWholeBoxplt <- ggplot(data = mydata, aes(y = WHOLE, x = CLASS, color=CLASS)) +
  geom_boxplot() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Whole Weight (grams)", color = "Class")
ggVolumeScatter <- ggplot(data = mydata, aes(y = VOLUME, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Class")
ggWholeScatter <- ggplot(data = mydata, aes(y = WHOLE, x = RINGS, color=CLASS)) +
  geom_point() +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Whole Weight (grams)", color = "Class")

grid.arrange(ggVolumeBoxplt, ggWholeBoxplt, ggVolumeScatter, ggWholeScatter, ncol=2)

```

**Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: There is a general positive trend between the number of rings and both the Whole and Volume variables. However, both variables have a wide range of variability for the A3, A4, and A5 classes. Therefore, I believe that both volume and whole would not perform well as predictors for age, unless the age range is limited to very young abelones.***

-----

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
mydataAgg <- aggregate(x = mydata[,c("VOLUME","SHUCK","RATIO")], by = list(mydata$SEX, mydata$CLASS), FUN = "mean")

# create volume matrix
volumeMatrix <- matrix(mydataAgg$VOLUME, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))
# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Volume by Sex and Class") %>% kable_styling()

# create shuck matrix
shuckMatrix <- matrix(mydataAgg$SHUCK, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))
# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Shuck Weight by Sex and Class") %>% kable_styling()

# create ratio matrix
ratioMatrix <- matrix(mydataAgg$RATIO, nrow = 3, ncol = 5, dimnames = list(unique(mydataAgg$Group.1), unique(mydataAgg$Group.2)))

# output to table
kable(volumeMatrix, format = "html", caption = "Abelone Ratio (Shuck Weight to Volume) by Sex and Class") %>% kable_styling()

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
# create ratio to class chart
ggplot(data = mydataAgg, aes(y = RATIO, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Ratio of Shuck Weight to Volume", color = "Sex", title = "Mean Ratio of Shuck Weight to Volume by Class")

# create volume to class chart
ggplot(data = mydataAgg, aes(y = VOLUME, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Volume (cm^3)", color = "Sex", title = "Mean Volume by Class")

# create shuck to class chart
ggplot(data = mydataAgg, aes(y = SHUCK, x = mydataAgg$Group.2, group = Group.1, color=Group.1)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  scale_color_brewer(palette="Dark2") +
  labs(x = "Class", y="Shuck Weight (grams)", color = "Sex", title = "Mean Shuck Weight by Class")

```

**Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
mydataSubset_i <- subset(mydata, RINGS < 16 & SEX == 'I')
mydataSubset_a <- subset(mydata, RINGS < 16 & (SEX == 'M' | SEX == 'F'))


par(mfrow = c(2, 2))
boxplot(VOLUME ~ RINGS, data = mydataSubset_i, col = rgb(44, 127, 184, maxColorValue = 255), ylim = c(0, 1100),
     main = "Rings of infants by Volume", ylab = "Volume")
boxplot(VOLUME ~ RINGS, data = mydataSubset_a, col = rgb(127, 205, 187, maxColorValue = 255), ylim = c(0, 1100),
     main = "Rings of adults by Volume", ylab = "Volume")
boxplot(WHOLE ~ RINGS, data = mydataSubset_i, col = rgb(44, 127, 184, maxColorValue = 255), ylim = c(0, 400),
     main = "Rings of infants by Whole Weight", ylab = "Whole Weight")
boxplot(WHOLE ~ RINGS, data = mydataSubset_a, col = rgb(127, 205, 187, maxColorValue = 255), ylim = c(0, 400),
     main = "Rings of adults by Whole Weight", ylab = "Whole Weight")


```

**Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (Enter your answer here.)***

-----

**Conclusions**

Please respond to each of the following questions (10 points total):

**Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Enter your answer here.)***

**Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (Enter your answer here.)***

**Question 3)  (2 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Enter your answer here.)***
