---
title: "Advanced Abalone Analysis"
author: "mcgovern, kevin"
subtitle: "Data Analysis #2 Version 2 (75 points total)"
output:
  html_document: default
---

Introduction paragraphs

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(flux)
library(ggplot2)
library(gridExtra)
library(moments)
# library(rockchalk) # base R code replaces requirement for this package
library(car)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",")
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```


###Section 1 (5 points)
(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a}

# build qq plot of ratio
q1.qqplot <- ggplot(data = mydata, aes(sample = RATIO)) +
  geom_qq(color='#2c7fb8') +
  geom_qq_line() +
  theme(legend.position = "none") +
  labs(x = "Theoretical", y = "Sample Ratio", title = "QQ Plot of Shuck Weight to Volume")

# build histogram of ratio
q1.hist <- ggplot(data = mydata, aes(x = RATIO)) +
  geom_histogram(binwidth=.02, fill='#2c7fb8') +
  theme(legend.position = "none") +
  labs(x = "Ratio", y = "Count", title = "Histogram of Shuck Weight to Volume")

grid.arrange(q1.qqplot, q1.hist, ncol=2)

# output skewness and kurtosis
print(paste("Skewness of ratio of shuck weight to volume", skewness(mydata$RATIO)))
print(paste("Kurtosis of ratio of shuck weight to volume", kurtosis(mydata$RATIO)))

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b}
# calculate metric for log ratio
mydata$L_RATIO <- log10(mydata$RATIO)

# build qq plot of log ratio
q1.qqplot <- ggplot(data = mydata, aes(sample = L_RATIO)) +
  geom_qq(color='#2c7fb8') +
  geom_qq_line() +
  theme(legend.position = "none") +
  labs(x = "Theoretical", y = "Sample Ratio", title = "QQ Plot of Log10 Ratio")

# build histogram of log ratio
q1.hist <- ggplot(data = mydata, aes(x = L_RATIO)) +
  geom_histogram(binwidth=.02, fill='#2c7fb8') +
  theme(legend.position = "none") +
  labs(x = "Ratio", y = "Count", title = "Histogram of Log10 Ratio")

grid.arrange(q1.qqplot, q1.hist, ncol=2)

# output skewness and kurtosis
print(paste("Skewness of ratio of shuck weight to volume", skewness(mydata$L_RATIO)))
print(paste("Kurtosis of ratio of shuck weight to volume", kurtosis(mydata$L_RATIO)))
```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
# Bartlett test of homogeneity across classes using ratio as dependent variable
print("Test raw ratio homogeneity")
bartlett.test(mydata$RATIO ~ mydata$CLASS)

# Bartlett test of homogeneity across classes using log ratio as dependent variable
print("Test log10 ratio homogeneity")
bartlett.test(mydata$L_RATIO ~ mydata$CLASS)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: (Enter your answer here.)***


###Section 2 (10 points)
(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}

# Two Way ANOVA with interaction term CLASS:SEX
q2.aov_interaction <- aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata)
summary(q2.aov_interaction)

# Two Way ANOVA without interaction term CLASS:SEX
q2.aov_wo_interaction <- aov(L_RATIO ~ CLASS + SEX, data = mydata)
summary(q2.aov_wo_interaction)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: (Enter your answer here.)***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(q2.aov_wo_interaction, conf.level = 0.95)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: Signif difference in means for log ratio of all classes except A1-A2, reject null that means are the same between classes and accept alternative. For sex, fail to reject the null that there is a significant difference between means of males and females. For difference in means between infants and males, as well as infants and females, reject null and accept alternative.***


###Section 3 (10 points)
(3)(a1) We combine "M" and "F" into a new level, "ADULT". (While this could be accomplished using *combineLevels()* from the 'rockchalk' package, we use base R code because many students do not have access to the rockchalk package.) This necessitated defining a new variable, TYPE, in mydata which had two levels:  "I" and "ADULT". 

```{r Part_3a1}
# here we show how to define the new variable TYPE using only base R functions (no need for outside packages)
mydata$TYPE <- character(nrow(mydata))  # initialize the TYPE column as all blanks
for (i in seq(along = mydata$SEX)) {
  mydata$TYPE[i] <- 'I'
  if (mydata$SEX[i] == 'M' || mydata$SEX[i] == 'F') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on definition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
table(mydata$SEX, mydata$TYPE)

```
(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2}

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: (Enter your answer here.)***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b}


```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: (Enter your answer here.)***


###Section 4 (5 points)
(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. Reclassify the infants in classes A4 and A5 as ADULTS. This reclassification could have been achieved using *combineLevels()*, but only on the abalones in classes A4 and A5. We will do this recoding of the TYPE variable using base R functions. We will use this recoded TYPE variable, in which the infants in A4 and A5 are reclassified as ADULTS, for the remainder of this data analysis assignment. 

```{r Part_4a1}
for (i in seq(along = mydata$TYPE)) {
  if (mydata$CLASS[i] == 'A4' || mydata$CLASS[i] == 'A5') mydata$TYPE[i] <- 'ADULT'
}
mydata$TYPE <- factor(mydata$TYPE)
cat('\nCheck on redefinition of TYPE object (should be an integer): ', typeof(mydata$TYPE))
cat('\nmydata$TYPE is treated as a factor: ', is.factor(mydata$TYPE), '\n')
cat('\nThree-way contingency table for SEX, CLASS, and TYPE:\n')
print(table(mydata$SEX, mydata$CLASS, mydata$TYPE))
```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}




```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: (Enter your answer here.)***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: (Enter your answer here.)***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

### Section 5 (5 points)
(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5a}


```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b}


```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: (Enter your answer here.)***

-----

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

###Section 6 (5 points)
(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

# prop.infants shows the impact of increasing the volume cutoff for
# harvesting. The following code shows how to "split" the population at
# a 50% harvest of infants.

n.infants <- sum(prop.infants <= 0.5)
split.infants <- min.v + (n.infants + 0.5)*delta  # This estimates the desired volume.
split.infants

n.adults <- sum(prop.adults <= 0.5)
split.adults <- min.v + (n.adults + 0.5)*delta
split.adults

```

(6)(b) Present a plot showing the infant proportions and the adult proportions versus volume.value. Compute the 50% "split" volume.value for each and show on the plot.   

```{r Part_6b}


```

**Essay Question:  The two 50% "split" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: (Enter your answer here.)***

-----

This part will address the determination of a volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

###Section 7 (10 points)
(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}


```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

```{r Part_7c}


```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions (NOTE:  the adult harvest proportion is the "true positive rate" and the infant harvest proportion is the "false positive rate").

Code for calculating the adult harvest proportion is provided.

```{r Part_7d}
(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

###Section 8 (10 points)
(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions.

```{r Part_8a}
volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

```

(8)(b) Another cutoff is one for which the proportion of adults not harvested equals the proportion of infants harvested. This cutoff would equate these rates; effectively, our two errors:  'missed' adults and wrongly-harvested infants. This leaves for discussion which is the greater loss:  a larger proportion of adults not harvested or infants harvested?  This cutoff is 237.7383. Calculate the separate harvest proportions for infants and adults using this cutoff. Show these proportions.  Code for determining this cutoff is provided.  

```{r Part_8b}
volume.value[which.min(abs(prop.adults - (1-prop.infants)))] # [1] 237.6391

```


###Section 9 (5 points)
(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (7) and (8) on this plot and label each. 

```{r Part_9}


```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}


```


###Section 10 (10 points)
(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population
 	
```{r Part_10} 	


```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the three cutoffs determined in this assignment.**   

***Answer: (Enter your answer here.)***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for implementation of a cutoff?  4)  What suggestions would you have for planning future abalone studies of this type? 

***Answer: (Enter your answer here.)***